---
title: "Mushroom Edibility Analysis"
author: "Tofu-FC - Huiwen Wang, Rocky Zhang, Darrick Zhang"
date: "10,28,24"
format: pdf
execute: 
  warning: false
  message: false
  echo: false
editor: visual
bibliography: references.bib
---

```{r}
#| label: load packages and data

library(tidyverse)
library(tidymodels)
library(dbplyr)
library(knitr)
library(GGally)
library(gridExtra)
library(caTools) # might need to install
library(caret) # might need to install
library(grid)
# add other packages as needed

mushrooms <- read.csv2("data/mushrooms_dataset.csv")

# Make Quantitative data numerical
mushrooms$cap.diameter <- as.numeric(mushrooms$cap.diameter)
mushrooms$stem.height <- as.numeric(mushrooms$stem.height)
mushrooms$stem.width <- as.numeric(mushrooms$stem.width)

# Convert the response variable to binary (e = 0, p = 1)
mushrooms <- mushrooms |>
  mutate(
    class_binary = ifelse(mushrooms$class == "e", 0, 1)
  )

# Simplified some categorical variables
mushrooms <- mushrooms |>
  mutate(
    stem.color.modified = case_when(
      stem.color %in% c("w", "y", "n") ~ stem.color, TRUE ~ "Other"),
    habitat.modified = case_when(
      habitat %in% c("d", "g") ~ habitat, TRUE ~ "Other")
  )

#Split data into train and test
set.seed(123)  # for reproducibility
split <- sample.split(mushrooms$class_binary, SplitRatio = 0.7)
train_data <- subset(mushrooms, split == TRUE)
test_data <- subset(mushrooms, split == FALSE)
```

## Introduction

### Project Motivation / Background:

Mushrooms are vital to the general wellness of the ecosystem, decomposing and recycling the nutrients in the soil. Mushrooms also provide a valuable food source full of nutrients for human beings and other important organisms. However, some mushroom species can also be poisonous and harmful.

The importance of this research has been highlighted in a multitude of studies. Take this quote, for example:

> The ingestion of wild and potentially toxic mushrooms is common in the United States, with poison centers logging cases in the National Poison Data System (NPDS) for over 30 years. From 1999 to 2016, there were 133,700 reported cases of mushroom exposure, mostly unintentional and involving children under six years old. While the majority of cases resulted in no or minor harm, there were 704 instances of major harm and 52 fatalities, primarily due to cyclopeptide-producing mushrooms ingested unintentionally by older adults. Misidentification of edible mushroom species is a common cause of poisoning and may be preventable through education [@brandenburgward2018].

As shown by studies and other similar studies, accurate classification of mushrooms is crucial for preventing poisoning incidents. Many toxic mushroom species closely resemble edible varieties, making it easy for foragers to misidentify them. Thus, our research will focus on what physical features and environmental factors of mushrooms humans can use to identify toxic/poisonous mushrooms in the wild. By conducting a research study on how to distinguish between safe and dangerous species, we can mitigate the incidence of mushroom poisoning and ensure safer foraging practices.

### Research Question:

What environmental factors and/or physical features of mushrooms indicate that a wild mushroom is poisonous?

### Hypothesis:

Mushrooms in the wild with obvious physical features like white gills, white rings, red caps, or red stems tend to be poisonous. These obvious physical traits are more likely to be spotted by animals, which would provide an evolutionary disadvantage unless they contain certain self-defense mechanisms, such as poison or toxins. Additionally, the habitat and season in which mushrooms are planted and grow may also affect whether they're poisonous. Different temperatures, humidity, and light can affect the production of toxins, which may also affect the edibility of mushrooms.

### Data Description:

The data was curated on April 26, 1987, and submitted to the UCI by the National Audubon Society Field Guide. The National Audubon Society conducted extensive field research throughout North America, recording their observations on various aspects of mushrooms. Their research incorporate a wide range of physical characteristics, including size, shape, color, and texture of the mushrooms. Additionally, they documented environmental factors such as the type of habitat and seasonal variations. Importantly, the study also focused on the toxicity of the mushrooms, noting which species were poisonous. This comprehensive dataset provides valuable insights into the relationship between mushrooms and their environments, contributing significantly to the understanding of the factors influencing mushroom toxicity.

Our response variable is `class`, which is a qualitative variable labeled "e" for edible or "p" for poisonous.

Because we want our classifier to be easily used by people, and quantitative predictors can be harder to measure, we will focus on only one. We are also interested in `cap.diameter`, the diameter of the mushroom cap (cm).

Key qualitative predictor variables include `cap.shape`, the shape of the mushroom cap; `gill.color`, the color of the fungi gills, `stem.color`, the color of the mushroom stem; `habitat`, the habitat that the mushroom is grown/found; and `season`, the season that the mushroom is grown/found (spring=s, summer=u, autumn=a, winter=w).

## Exploratory Data Analysis

```{r}
#| label: check rows

# # Count the number of rows with at least one missing value
# num_missing_rows <- sum(rowSums(is.na(mushrooms)) > 0)
# cat("Number of rows with at least one missing value:", num_missing_rows, "\n")
```

```{r}
mushrooms |>
  count(class) |>
  mutate(percentage = round(n / sum(n),3)) |>
  kable(digits=3)
```

```{r, eval=FALSE}
#| label: univariate response variable, summary statistics

#EDA Response (Categorical) Variable
ggplot(mushrooms, aes(x = class, fill = class)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Distribution of Edibility/Classes of Mushrooms",
    x = "Class",
    caption = "44.5% are edible, 55.5% are poisonous"
  ) +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5)

# Summary Statistics
# mushrooms |>
#  group_by(class) |>
#  summarise(
#    count = n(),
#    percentage = (count / nrow(mushrooms)) * 100
#  ) |> kable(digits=3)
```

Looking at the overall distribution of our response variable `class`, most of the mushrooms in our dataset seem to be poisonous ("p"). 33888 of the observations, or 55.5% of them are labeled poisonous, as opposed to 27181 (44.5%) of them as edible.

```{r, eval=FALSE}
#| label: multicolinearity

mushrooms |>
  select(cap.diameter, stem.height, stem.width) |>
  ggpairs()

mushrooms |>
  summarise(min = min(cap.diameter), 
            q1 = quantile(cap.diameter, 0.25), 
            median = median(cap.diameter),
            q3 = quantile(cap.diameter, 0.75), 
            max = max(cap.diameter),
            mean = mean(cap.diameter),
            sd = sd(cap.diameter)) |>
  kable(digits = 3, 
        caption = "cap diameter summary statistics")

mushrooms |>
  summarise(min = min(stem.height), 
            q1 = quantile(stem.height, 0.25), 
            median = median(stem.height),
            q3 = quantile(stem.height, 0.75), 
            max = max(stem.height),
            mean = mean(stem.height),
            sd = sd(stem.height)) |>
  kable(digits = 3,
        caption = "stem height summary statistics")

mushrooms |>
  summarise(min = min(stem.width), 
            q1 = quantile(stem.width, 0.25), 
            median = median(stem.width),
            q3 = quantile(stem.width, 0.75), 
            max = max(stem.width),
            mean = mean(stem.width),
            sd = sd(stem.width)) |>
  kable(digits = 3,
        caption = "stem width summary statistics")

# OLD Narrative:
# We were first interested in seeing the relationship between our continuous, quantitative variables. While not totally linear, with correlations of $0.423$ and $0.436$, they graphs seem to be somewhat linear, with some redundancy in their information. Thus, for the rest of the EDA we will focus mainly on visualizing cap.diameter. We may consider adding them back for the final model, but were more interested in seeing some of the EDA with the categorical variables. The mean cap diameter is 6.734cm, with a SD of 5.265cm. The mean stem height is 6.582cm, with a SD of 3.37cm. The mean stem width is 12.149cm, with a SD of 10.036cm.
```

```{r,fig.width=8, fig.height=4}
#| label: univariate predictor

# mushrooms |>
#   ggplot(aes(x = cap.diameter)) + 
#   geom_density(fill = "skyblue", alpha = 0.7) +
#   labs(title = "Kernel Density Plot of cap diameter",
#        x = "Cap diameter (cm)",
#        y = "Density")

ggplot(mushrooms, aes(x = cap.diameter)) +
    geom_histogram(fill = "skyblue", alpha = 0.9) +
  geom_vline(xintercept=mean(mushrooms$cap.diameter), color="red") +
    theme_minimal() +
    labs(title = "Cap diameter",
         x="Cap diameter")

mushrooms |>
  summarise(min = min(cap.diameter), 
            q1 = quantile(cap.diameter, 0.25), 
            median = median(cap.diameter),
            q3 = quantile(cap.diameter, 0.75), 
            max = max(cap.diameter),
            mean = mean(cap.diameter),
            sd = sd(cap.diameter)) |>
  kable(digits = 3, 
        caption = "cap diameter summary statistics")
```

Visualizing the shape of our quantitative predictor, cap diameter, the distribution seems to be roughly unimodal, skewed right. The mean cap diameter is 6.734cm, with a standard deviation of 5.265cm.

Since the rest of our predictors our qualitative, we report their distributions through the tables below:

| g       | l      | m       | p     | h      | u     | w     | d     |
|---------|--------|---------|-------|--------|-------|-------|-------|
| grasses | leaves | meadows | paths | heaths | urban | waste | woods |

```{r}
capshapecodes <- c("bell", "conical", "flat", "other", "spherical", "sunken", "convex")
capcolorcodes <- c("buff", "red", "gray", "black", "blue", "brown", "orange", "pink", "green", "purple", "white", "yellow")
stemcolorcodes <- c("buff", "red", "none", "gray", "black", "blue", "brown", "orange", "pink", "green", "purple", "white", "yellow")
gillcolorcodes <- c("buff", "red", "none", "gray", "black", "brown", "orange", "pink", "green", "purple", "white", "yellow")
habitatcodes <- c("woods", "grasses", "heaths", "leaves", "meadows", "paths", "urban", "waste")
seasoncodes <- c("autumn", "spring", "summer", "winter") 

temp_data <- mushrooms |>
  count(cap.shape) |>
  mutate(percentage = round(n / sum(n),3))

rotated_data <- as.data.frame(t(temp_data[, -2]))
# colnames(rotated_data) = rotated_data[1,]
colnames(rotated_data) = capshapecodes
rownames(rotated_data) = rownames(rotated_data)
rotated_data |> kable(digits=3)

temp_data <- mushrooms |>
  count(cap.color) |>
  mutate(percentage = round(n / sum(n),3))

rotated_data <- as.data.frame(t(temp_data[, -2]))  
rownames(rotated_data) = rownames(rotated_data)
colnames(rotated_data) = capcolorcodes
rotated_data |> kable(digits=3)

temp_data <- mushrooms |>
  count(habitat) |>
  mutate(percentage = round(n / sum(n),3))

rotated_data <- as.data.frame(t(temp_data[, -2]))
rownames(rotated_data) = rownames(rotated_data)
colnames(rotated_data) = habitatcodes
rotated_data |> kable(digits=3)


temp_data <- mushrooms |>
  count(season) |>
  mutate(percentage = round(n / sum(n),3))

rotated_data <- as.data.frame(t(temp_data[, -2]))
colnames(rotated_data) = seasoncodes
rownames(rotated_data) = rownames(rotated_data)
rotated_data |> kable(digits=3)


temp_data <- mushrooms |>
  count(stem.color) |>
  mutate(percentage = round(n / sum(n),3))

rotated_data <- as.data.frame(t(temp_data[, -2]))
colnames(rotated_data) = stemcolorcodes
rownames(rotated_data) = rownames(rotated_data)
rotated_data |> kable(digits=3)


temp_data <- mushrooms |>
  count(gill.color) |>
  mutate(percentage = round(n / sum(n),3))

rotated_data <- as.data.frame(t(temp_data[, -2]))
colnames(rotated_data) = gillcolorcodes
rownames(rotated_data) = rownames(rotated_data)
rotated_data |> kable(digits=3)

```

For qualitative variables, there appears to be more common physical and environmental characteristics. For example, for cap shape, flat and convex tends to be the most common; for stem color the most common is white, yellow, and brown; for habitat, woods is the most common. Thus, there are also characteristics which happen to be rarer, yet for some reason, natural selection has decided to preserve. These characteristics may have evolutionary advantageous properties (such as being poisonous), and we hope that they help us in our logistic regression model.

```{r, fig.width=8, fig.height=8}
#| label: bivariate EDA
g1 <- ggplot(mushrooms, aes(x = cap.diameter, color = class)) +
    geom_boxplot() +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs cap diameter",
         x = "Cap diameter")
g2 <- ggplot(mushrooms, aes(x = cap.shape, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs cap shape",
         x = "Cap shape")
g3 <- ggplot(mushrooms, aes(x = gill.color, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs gill color",
         x = "Gill color")
# g3 <- ggplot(mushrooms, aes(x = stem.root, fill = class)) +
#     geom_bar(position = "dodge") +
#     labs(title = "Response vs stem root")
g4 <- ggplot(mushrooms, aes(x = stem.color, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs stem color",
         x ="Stem color")
g5 <- ggplot(mushrooms, aes(x = habitat, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs habitat",
         x = "Habitat")
g6 <- ggplot(mushrooms, aes(x = season, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs season",
         x = "Season")

# grid.arrange(g1,g2,g3,g4,g5,g6,bottom = textGrob("cap shape: bell=b, conical=c, convex=x, flat=f, sunken=s, spherical=p, others=o
# colors: brown=n, buff=b, gray=g, green=r, pink=p, purple=u, red=e, white=w, yellow=y, blue=l, orange=o, black=k
# habitat: grasses=g, leaves=l, meadows=m, paths=p, heaths=h, urban=u, waste=w, woods=d
# season: spring=s, summer=u, autumn=a, winter=w",
#                                                  gp = gpar(fontsize = 8)))

grid.arrange(g1,g2,g3,g4,g5,g6)
```

Looking at this bivariate exploratory data analysis, we see that, on average, smaller cap diameters seem to be correlated with poisonous mushrooms. We also observe some categories with a large disparity between the number of edible and number of poisonous mushrooms, offering potential modeling power. For example, if we randomly select a mushroom with a cap shape of convex, bell, or others, it is more likely to be poisonous/toxic than edible. Similarly, we see that mushrooms with gill color of brown and yellow tends to be poisonous. However, in many cases, it is hard to accurately predict whether a mushroom is edible or poisonous based off just one characteristic, suggesting our model needs to incorporate multiple predictors and/or interaction terms.

One interaction term we were interested in looking at is `habitat*season`. Mushrooms that grow in the same habitat may have different toxicity classification depending on if being poisonous is needed to defend against predators. The number of predators themselves may vary depending on season, so season may change how habitat affects the log-odds of whether the mushroom is edible or poisonous.

```{r}
#| label: analysis of interaction

ggplot(mushrooms, aes(x = habitat, fill = class)) +
  geom_bar(position = "fill") + 
  facet_wrap(~ season) +
  labs(title = "Edibility Distribution by Habitat and Season",
       x = "Habitat",
       y = "Proportion of Edibility",
       caption = "Habitat codes: grasses=g, leaves=l, meadows=m, paths=p, heaths=h, urban=u, waste=w, woods=d
       Season codes: spring=s, summer=u, autumn=a, winter=w") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Looking at this graph, we see that mushrooms in certain habitats might only be edible during specific seasons. For example, mushrooms in the meadows are edible in the winter, but may be poisonous in other seasons. This suggests we may want to incorporate this interaction term into our final model.

We were also interested in looking at the interaction between cap color and cap shape, as these are two of the characteristics which are most apparent to a potential predator.

```{r}
#| label: heatmap

mushroom_prop <- mushrooms |>
  group_by(cap.color, cap.shape) |> 
  summarize(edible_count = sum(class == "e"),
            poisonous_count = sum(class == "p")) |>
  mutate(proportion_e_to_p = edible_count / poisonous_count) |>
  select(cap.color, cap.shape, proportion_e_to_p)
mushroom_prop <- mushroom_prop |>
  mutate(log_prop = log(proportion_e_to_p))



ggplot(mushroom_prop, aes(x = cap.color, y = cap.shape, 
                          fill = log_prop)) +
  geom_tile() +
  labs(title = "Cap Color and Shape Combinations by Edibility",
       subtitle = "Using log ratios of edible to poisonous mushrooms",
       x = "Cap Color",
       y = "Cap Shape",
       fill = "log(e/p)") +
  geom_tile(data = mushroom_prop |> filter(log_prop == -Inf),
            aes(y = cap.shape, x = cap.color),
            fill = "firebrick1") +
  geom_tile(data = mushroom_prop |> filter(log_prop == Inf),
            aes(y = cap.shape, x = cap.color),
            fill = "darkseagreen2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  geom_text(aes(label = case_when(
                          log_prop == Inf ~ "e",
                          log_prop == -Inf ~ "p",
                          TRUE ~ as.character(round(log_prop, 1))),
                color = case_when(
                          log_prop == Inf ~ "always edible",
                          log_prop == -Inf ~ "always poisonous",
                          TRUE ~ "other")),
            show.legend = TRUE) +
  scale_color_manual(values = c("always edible" = "darkseagreen2", 
                                "always poisonous" = "firebrick1", 
                                "other" = "black"),
                     name = "Edibility")
```

As the heatmap shows, certain combinations of cap color and cap shape are always edible or poisonous. Additionally, the log ratios across combinations of cap colors and cap sizes are varied with no pattern – for a mushroom with a sunken cap shape, it could be always edible (if the color is buff) to always poisonous (if the color is red). Similarly, if a mushroom is brown, it could be high likely it is edible (if the cap shape is conical) or likely it is poisonous (if the cap shape is bell). Thus, we may have to consider this interaction effect in our final model.

## Analysis

The base model:

$$
\log\left(\frac{\text{P}(\text{class = poisonous})}{\text{P}(\text{class = edible})}\right) = \beta_0 + \beta_1 \cdot \text{cap.diameter} + \beta_2 \cdot \text{season} + \beta_3 \cdot \text{cap.shape} + \beta_4 \cdot \text{cap.color} \\+ \beta_5 \cdot \text{gill.color} + \beta_6 \cdot \text{stem.color} + \beta_7 \cdot \text{habitat}
$$Predictor terms were chosen from the EDA and general physical or environmental factors that are generally understood and easy to evaluate by everyone. To determine if any predictors may not useful, we looked at coefficients from the tidy function with p-values greater than 0.01

```{r}
#| label: Tidy Table
main_model <- glm(class_binary ~ cap.diameter + season + cap.shape + cap.color +
                    gill.color + stem.color + habitat, 
                    data = mushrooms, family = "binomial")
main_model |>
  tidy() |>
  filter(p.value > 0.01) |>
  kable(digits = 3)
```

The Wald's Significance Tests for coefficients of multiple categories of the same predictor variables reveals that for certain categories there may be limited data (also seen through EDA) and/or limited predictive power. For simplicity of our model, we combine these categories into a general "Other" category. For example, `stem.color` of "w", "y", and "n" were kept while the other observations were assigned to a general "Other" category. For habitat, "d" and "g" were kept.

Running a likelihood ratio test to evaluate the overall significance of the coefficents of the new model with modified categorical variables, we have:

```{r}
#| label: Overall Significance F Test Redone with Modified Categoricals (Likelihood Ratio Test)
null_model <- glm(class_binary ~ 1, data = mushrooms, family = "binomial")
main_model_modified <- glm(class_binary ~ cap.diameter + season + cap.shape + cap.color +
                    gill.color + stem.color.modified + habitat.modified, 
                    data = mushrooms, family = "binomial")
#(L_0 <- glance(null_model)$logLik)
#(L_a <- glance(main_model_modified)$logLik)
#(G <- -2 * (L_0 - L_a)) #LRT test statistic
#(p_value <- pchisq(G, df = 3, lower.tail = FALSE))
anova(null_model, main_model_modified, test = "Chisq") |> 
  tidy() |> 
  select(-df.residual) |>
  kable(digits = 3)
```

$$
H_0: \beta_j = 0 \\ H_a: \beta_j \neq 0 \ for\ at\ least\ 1\ j
$$

Since the p-value is small, and less than $\alpha = 0.05$, we reject the $H_0$. The data provide sufficent evidence of at least one non-zero coefficient in the model.

The model coefficients and corresponding inferential statistics are shown below:

```{r}
main_model_modified |>
  tidy() |>
  kable(digits = 3)
```

As shown on our EDA, we hypothesized there may be some potential interaction terms. To determine the need for them in our model, we performed a drop in deviance test with the added interaction terms of `habitat*season` and `cap.shape*cap.color`.

```{r, eval=FALSE}
#| label: comments/to-dos

#High p-values in tidy table of full_model? Is that a point of concern? The likelihood test is not picking up on stem color? Are the p-values for 1 term the wald's test? Intercept p value is super high, should we exclude it? #Testing anova with the full dataset not just the train dataset

#EDA univariate or bivariate to make it shorter and more readable?

#Make height of the bivariate taller, and describes trends, don't really need univariate EDA. Simplify categories of some predictors to change with p-values. All the modeling should be done with the training set. Prediction- AUC, BIC as priority. Add more to the tables so we can remove univariate eda.
```

```{r}
#| label: Drop-in-deviance Test

full_model <- glm(class_binary ~ cap.diameter + season + cap.shape + cap.color +
                    gill.color + stem.color.modified + habitat.modified + habitat.modified*season + 
                    cap.shape*cap.color,
                    data = mushrooms, family = "binomial")

anova(main_model_modified, full_model, test = "Chisq") |> 
  tidy() |> 
  select(-df.residual) |>
  kable(digits = 3)
```

Since the p-value is low below $\alpha = 0.05$, we decide to include these interaction terms as there is convincing evidence that at least one of these interactive term coefficients are not 0 and thus helpful in the model.

Additionally for the base model, the AIC is `{r} glance(main_model_modified)$AIC |> signif(digits=3)` and the BIC is `{r} glance(main_model_modified)$BIC |> signif(digits=3)`, whereas for the model with interaction effects, the AIC is `{r} glance(full_model)$AIC |> signif(digits=3)` and the BIC is `{r} glance(full_model)$BIC |> signif(digits=3)`. For both measures, the full model performs better (lower AIC/BIC).

## Model Results

### Final Model

$$
\log(\text{Odds(class = poisonous}) = \beta_0 + \beta_1 \cdot \text{cap.diameter} + \beta_2 \cdot \text{season} + \beta_3 \cdot \text{cap.shape} \\ 
+
 \beta_4 \cdot \text{cap.color} + \beta_5 \cdot \text{gill.color} + \beta_6 \cdot \text{stem.color.modified} + \beta_7 \cdot \text{habitat.modified} + \beta_8 \cdot (\text{habitat.modified} \cdot \text{season}) \\
 + \beta_9 \cdot (\text{cap.shape} \cdot \text{cap.color})
$$

### ROC curve

```{r}
test_data$class_binary <- as.factor(test_data$class_binary)
predicted_values <- predict(full_model, test_data, type = "response")
test_data <- test_data |> 
  mutate(pred_prob = predicted_values)

roc_curve_data <- test_data |>
  roc_curve(class_binary, pred_prob, event_level = "second")
autoplot(roc_curve_data)

test_data_aug <- augment(full_model, newdata = test_data)
test_data_aug$class_binary <- as.factor(test_data_aug$class_binary)
test_data_aug |>
  roc_auc(class_binary, .fitted, event_level = "second")
```

```{r}
roc_curve_data |>
  filter(sensitivity <= 0.9501) |>
  head(n=2)

threshold <- 0.247
test_data <- test_data |> 
  mutate(
    pred_class = case_when(
      pred_prob < threshold ~ 0,
      TRUE ~ 1
    )
  )
```

The model is decent as the AUC is 0.797 which is closer to 1 than 0.5. We decided on a threshold of p = 0.247 to achieve a sensitivity of 95%, since we wanted to prioritize minimizing false negatives, which are more expensive – better to be careful than eat a poisonous mushroom classified as "edible".

\$\$

```{r}
#| label: confusion matrix

test_data$pred_class <- as.factor(test_data$pred_class)
conf_mat <- test_data |>
  yardstick::conf_mat(class_binary, pred_class)
autoplot(conf_mat, type = "heatmap") +
  labs(
    title = "Confusion matrix with p=0.247",
    subtitle = "0: edible; 1: poisonous"
  )
```

Using this threshold, we can evaluate our model's performance with a confusion matrix. As desired, for poisonous mushrooms, we are able to successfully classify 95% of them as poisonous. Our model struggles at correctly identifying mushrooms which are actually edible as edible, with a false positive rate of 73.7%. We were able to build a model that overall does much better (see appendix). However, this model requires many more variables, and becomes much more complex. For the purposes of this model, we wanted it to be applicable even in situations where humans found themselves having to assess edibility without many special tools or knowledge.

## References

## Appendix

```{r}
#| label: univariate predictors

p1 <- ggplot(mushrooms, aes(x = cap.diameter)) +
    geom_histogram() +
  geom_vline(xintercept=mean(mushrooms$cap.diameter), color="red") +
    theme_minimal() +
    labs(title = "Cap diameter",
         x="Cap diameter",
         caption = "mean: 6.734; SD: 5.265")
p2 <- ggplot(mushrooms, aes(x = cap.shape)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Cap shape",
         x="Cap shape")
p3 <- ggplot(mushrooms, aes(x = gill.color)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Gill color",
         x="Gill color")
p4 <- ggplot(mushrooms, aes(x = stem.color)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Stem color",
         x="Stem color")
p5 <- ggplot(mushrooms, aes(x = habitat)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Habitat",
         x="Habitat")
p6 <- ggplot(mushrooms, aes(x = season)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Season",
         x="Season")
grid.arrange(p1,p2,p3,p4,p5,p6,
              bottom = textGrob("cap shape: bell=b, conical=c, convex=x, flat=f, sunken=s, spherical=p, others=o
colors: brown=n, buff=b, gray=g, green=r, pink=p, purple=u, red=e, white=w, yellow=y, blue=l, orange=o, black=k
habitat: grasses=g, leaves=l, meadows=m, paths=p, heaths=h, urban=u, waste=w, woods=d
season: spring=s, summer=u, autumn=a, winter=w", 
                                                 gp = gpar(fontsize = 8)))
```

Here is our original univariate EDA with the full visualizations. As

```{r}
#| label: multivariate EDA
ggplot(mushrooms, aes(y = stem.height, x = stem.width, color = class)) +
  geom_point() +
  labs(
    title = "Distribution of Stem Height vs. Stem Width Among Different Edibility Classes",
    y = "Stem Height (cm)",
    x = "Stem Width (cm)",
    color = "Class (Edibile or Not)"
  ) +
  theme_minimal()
```

Here, we look at multivariate exploratory data analysis including 2 predictors and our response variable. We visualize the effect of both stem width and stem height on the response variable, class. Interestingly, it seems like mushrooms with either high stem width or stem height seem to be edible. This suggests there may be some potential interaction effects between stem height and stem width – the low value of one alone does not seem to predict if the mushroom is poisonous, but requires the low value of both. However, in our model when we added this interaction effect, the performance did not include that drastically, and we deemed it more important to keep the model parsimonious as possible. Additionally, quantitative features can be hard to measure, and so may be less practical when serving as a general guideline for foraging mushrooms.

```{r}
#| label: more interaction effects/vars?

full_model2 <- glm(class_binary ~ cap.diameter + season + cap.shape + cap.color +
                    gill.color + habitat + stem.root*stem.color + veil.type + veil.color +
                     has.ring + ring.type +
                     cap.shape*cap.color + habitat*season,
                   data = mushrooms, family="binomial")
full_model3 <- glm(class_binary ~ cap.diameter + season + cap.shape + cap.color +
                     gill.color + habitat + stem.root*stem.color + veil.type + veil.color + 
                     has.ring*ring.type +gill.attachment*gill.spacing+
                     cap.shape*cap.color + habitat*season,
                   data = mushrooms, family="binomial")
anova(full_model2, full_model3, test = "Chisq") |> 
  tidy() |> 
  kable(digits = 3)

glance(full_model2)$AIC
glance(full_model3)$AIC
glance(full_model2)$BIC
glance(full_model3)$BIC
```

Here, we played with adding more predictors to our model. We do achieve better ROC curves with these as well, but we decided that a smaller model would still be better, and that many of these predictors that we added here may be hard to identify for the average person.
