---
title: "Mushroom Edibility Analysis"
author: "Tofu FC: Huiwen Wang, Rocky Zhang, Darrick Zhang"
date: "2024-11-11"
date-format: "MMM DD, YYYY"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  warning: false
  message: false
  echo: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| label: load packages and data

library(tidyverse)
library(tidymodels)
library(dbplyr)
library(knitr)
library(GGally)
library(gridExtra)
library(caTools) # might need to install
library(caret) # might need to install
# add other packages as needed

mushrooms <- read.csv2("../data/mushrooms_dataset.csv")

# Make Quantitative data numerical
mushrooms$cap.diameter <- as.numeric(mushrooms$cap.diameter)
mushrooms$stem.height <- as.numeric(mushrooms$stem.height)
mushrooms$stem.width <- as.numeric(mushrooms$stem.width)
```

## Background

::: incremental
-   Mushrooms important to the ecosystem, serve as food source

-   But mushrooms can be potentially poisonous

    -   1999 to 2016, 133,700 reported cases

    -   704 instances of major harm and 52 fatalities

-   Can help reduce accidents by classifying poisonous mushrooms!

-   What environmental factors and/or physical features of mushrooms indicate that a wild mushroom is poisonous?
:::

#### 

## Data

::: incremental
-   Curated on April 26, 1987, and submitted to the UCI by the National Audubon Society Field Guide.
-   The National Audubon Society conducted extensive field research throughout North America, recording their observations on various aspects of mushrooms.
-   Their research incorporate a wide range of physical characteristics, including size, shape, color, and texture of the mushrooms.
-   Additionally, they documented environmental factors such as the type of habitat and seasonal variations along with the toxicity of the mushrooms.
:::

## EDA

::: panel-tabset
### Response Var.

```{r}
#EDA Response (Categorical) Variable
ggplot(mushrooms, aes(x = class, fill = class)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Distribution of Edibility/Classes of Mushrooms",
    x = "Class",
    caption = "44.5% are edible, 55.5% are poisonous"
  ) +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5)
```

### Univariate predictors

```{r}
#| label: univariate predictors

p1 <- ggplot(mushrooms, aes(x = cap.diameter)) +
    geom_histogram() +
    theme_minimal() +
    labs(title = "Cap diameter",
         x="Cap diameter")
p2 <- ggplot(mushrooms, aes(x = cap.shape)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Cap shape",
         x="Cap shape")
p3 <- ggplot(mushrooms, aes(x = gill.color)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Gill color",
         x="Gill color")
p4 <- ggplot(mushrooms, aes(x = stem.color)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Stem color",
         x="Stem color")
p5 <- ggplot(mushrooms, aes(x = habitat)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Habitat",
         x="Habitat")
p6 <- ggplot(mushrooms, aes(x = season)) +
    geom_bar() +
    theme_minimal() +
    labs(title = "Season",
         x="Season")
grid.arrange(p1,p2,p3,p4,p5,p6)
```

### Bivariate predictors

```{r}
#| label: bivariate EDA

g1 <- ggplot(mushrooms, aes(x = cap.diameter, color = class)) +
    geom_boxplot() +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs cap diameter",
         x = "Cap diameter")
g2 <- ggplot(mushrooms, aes(x = cap.shape, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs cap shape",
         x = "Cap shape")
g3 <- ggplot(mushrooms, aes(x = gill.color, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs gill color",
         x = "Gill color")
# g3 <- ggplot(mushrooms, aes(x = stem.root, fill = class)) +
#     geom_bar(position = "dodge") +
#     labs(title = "Response vs stem root")
g4 <- ggplot(mushrooms, aes(x = stem.color, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs stem color",
         x ="Stem color")
g5 <- ggplot(mushrooms, aes(x = habitat, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs habitat",
         x = "Habitat")
g6 <- ggplot(mushrooms, aes(x = season, fill = class)) +
    geom_bar(position = "dodge") +
    theme(legend.title = element_blank()) +
    labs(title = "Response vs season",
         x = "Season")

grid.arrange(g1,g2,g3,g4,g5,g6)
```
:::

## EDA (cont.)

::: panel-tabset
### Habitat and Season

```{r}
ggplot(mushrooms, aes(x = habitat, fill = class)) +
  geom_bar(position = "fill") + 
  facet_wrap(~ season) +
  labs(title = "Edibility Distribution by Habitat and Season",
       x = "Habitat",
       y = "Proportion of Edibility",
       caption = "Habitat codes: grasses=g, leaves=l, meadows=m, paths=p, heaths=h, urban=u, waste=w, woods=d") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Cap Color and Size

```{r}
mushroom_prop <- mushrooms |>
  group_by(cap.color, cap.shape) |> 
  summarize(edible_count = sum(class == "e"),
            poisonous_count = sum(class == "p")) |>
  mutate(proportion_e_to_p = edible_count / poisonous_count) |>
  select(cap.color, cap.shape, proportion_e_to_p)
mushroom_prop <- mushroom_prop |>
  mutate(log_prop = log(proportion_e_to_p))

caption_msg <- "'e' denotes always edible
                'p' denotes always poisonous
                Cap shape codes: bell=b, conical=c, convex=x, flat=f, sunken=s, spherical=p, others=o
                Cap color codes: brown=n, buff=b, gray=g, green=r, pink=p, purple=u, red=e, white=w, yellow=y, blue=l, orange=o, black=k"


ggplot(mushroom_prop, aes(x = cap.color, y = cap.shape, 
                          fill = log_prop)) +
  geom_tile() +
  labs(title = "Cap Color and Shape Combinations by Edibility",
       subtitle = 
         "Using log ratios of edible to poisonous mushrooms",
       x = "Cap Color",
       y = "Cap Shape",
       fill = "log(e/p)",
       caption = caption_msg) +
  geom_tile(data = mushroom_prop |>
              filter(log_prop == -Inf),
            aes(y = cap.shape, x = cap.color),
            fill = "firebrick1") +
  geom_tile(data = mushroom_prop |>
              filter(log_prop == Inf),
            aes(y = cap.shape, x = cap.color),
            fill = "darkseagreen2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  geom_text(aes(label = case_when(
                          log_prop == Inf ~ "e",
                          log_prop == -Inf ~ "p",
                          TRUE ~ as.character(round(log_prop, 1)))),
            color = "black")
```
:::

## Initial modeling

::: panel-tabset
### Simple Model

$$
\text{logit}(Odds(\text{class} = p)) = \beta_0 + \beta_1 \cdot \text{cap.diameter} + \beta_2 \cdot \text{season}
$$

### Tidy Output

```{r}
# Convert the response variable to binary (e = 0, p = 1)
mushrooms <- mushrooms |>
  mutate(
    class_binary = ifelse(mushrooms$class == "e", 0, 1)
  )

set.seed(123)  # for reproducibility
split <- sample.split(mushrooms$class_binary, SplitRatio = 0.7)
train_data <- subset(mushrooms, split == TRUE)
test_data <- subset(mushrooms, split == FALSE)

model <- glm(class_binary ~ cap.diameter + season, data = train_data, family = binomial)

tidy(model) |>
  kable(digits = 3)
```

### Confusion Matrix

class binary: edible = 0, poisonous = 1

```{r}
# Make predictions
predicted_probs <- predict(model, test_data, type = "response")
predicted_classes <- ifelse(predicted_probs > 0.5, 1, 0) # cutoff is p = 0.5

# Evaluate model
confusionMatrix(as.factor(predicted_classes), as.factor(test_data$class_binary))
```
:::

## Next steps + questions
::: incremental
-   Determine the need of interaction terms
-   Test more specified models
-   Evaluate an efficient parsimonious model
-   Optimize on p cutoff
-   Sensitivity vs. specificity
:::
